---
title: "COVID19"
author: "RQ"
date: "21/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(lubridate)
library(plotly)
library(zoo)
```

Thierry downloaded NM flight table. Stored on P and then moved to project folder, data-src.

Load NM flight table for data preparation.

```{r}
fn <- "./data-src/T_FLIGHT 01012020-15012020.xlsx"
sh <- excel_sheets(fn)
ds <- readxl::read_excel(fn, sheet = "Sheet 1", col_names = TRUE)
```

names(ds)

flight table vars:
[1] "FLT_UID"                     "LOBT"                        "ID"                         
 [4] "TACT_ID"                     "AIRCRAFT_ID"                 "REGISTRATION"               
 [7] "AIRCRAFT_TYPE_ICAO_ID"       "AIRCRAFT_OPERATOR"           "ADEP"                       
[10] "ADES"                        "ST_DEP"                      "ST_DES"                     
[13] "AIRAC_CYCLE"                 "IOBT"                        "EOBT_1"                     
[16] "ARVT_1"                      "FL_TOP_1"                    "RTE_LEN_1"                  
[19] "TAXI_TIME_1"                 "FLT_DUR_1"                   "COBT_2"                     
[22] "ARVT_2"                      "FL_TOP_2"                    "RTE_LEN_2"                  
[25] "TAXI_TIME_2"                 "FLT_DUR_2"                   "ATFM_DELAY"                 
[28] "MOST_PEN_REG"                "REG_ID"                      "REG_DAY"                    
[31] "REG_PERIOD"                  "REG_LOC"                     "REG_REASON"                 
[34] "IATA_DELAY_CODE"             "TV_ID"                       "AOBT_3"                     
[37] "ARVT_3"                      "FL_TOP_3"                    "RTE_LEN_3"                  
[40] "TAXI_TIME_3"                 "FLT_DUR_3"                   "ICAO_EQPT"                  
[43] "DATA_LINK_EQPT"              "COM_EQPT"                    "NAV_EQPT"                   
[46] "SSR_EQPT"                    "SURVIVAL_EQPT"               "PERFORMANCE"                
[49] "WK_TBL_CAT"                  "FUEL_ENDURANCE"              "PERSONS_ON_BOARD"           
[52] "ICAO_FLT_TYPE"               "NORTH_ATLANTIC"              "SENSITIVE"                  
[55] "FLT_STATUS"                  "FLT_STATUS_BFR_ACT"          "EXMP_RSN"                   
[58] "EXMP_RSN_LH"                 "FLT_PLN_SOURCE"              "FLT_DATA_QUAL"              
[61] "FLT_DATA_QUAL_ORIG"          "LATE_UPDATER"                "LATE_FILER"                 
[64] "FLT_RULES"                   "MAN_EXMP"                    "MAN_EXMPT"                  
[67] "REROUTING_STATE"             "SLOT_FORCED"                 "SSR_CODE"                   
[70] "SUSP_STATUS"                 "PERF_ROLE"                   "IFPS_PLAN_ID"               
[73] "RVR_VAL"                     "REROUTING_REASON"            "SELCAL_CODE"                
[76] "PREDEF_SNSTV_HDL_RSN"        "FL_REQ"                      "NR_REG_AFF"                 
[79] "NR_REG_EXL"                  "OPERATING_AIRCRAFT_OPERATOR" "AIRCRAFT_ADDRESS_SOURCE"    
[82] "AIRCRAFT_ADDRESS"            "IFPZ_ENTRY_TIME_EST"         "IFPZ_EXIT_TIME_EST"         
[85] "IFPZ_ENTRY_TIME_ACT"         "IFPZ_EXIT_TIME_ACT"          "DATA_LINK_TYPE"             
[88] "FPL_REMARK"                  "SNSTV_HDL_RSN"               "ADES_FILED"                 



```{r}
cols <- c("FLT_UID","LOBT","AIRCRAFT_ID","REGISTRATION","AIRCRAFT_TYPE_ICAO_ID","AIRCRAFT_OPERATOR","ADEP","ADES","ADES_FILED","AOBT_3","ARVT_3","TAXI_TIME_3","RTE_LEN_3","FLT_DUR_3","WK_TBL_CAT","ICAO_FLT_TYPE","OPERATING_AIRCRAFT_OPERATOR","AIRCRAFT_ADDRESS")

ds2 <- ds %>% select(cols)

# write_csv(ds2, "./data/2020-01-I.csv")
# ds <- read_csv("./data/2020-01-I.csv")
```

combine with airport data from ourairports.com

```{r}
url_ourapts <- "https://ourairports.com/data/airports.csv"
# apt_info    <- read_csv(url_ourapts)
# write_csv(apt_info, "./data/apts_ourairports.csv")

# extract relevant columns and clean names
# filter for "regular" ICAO location indicators
apt_info <- apt_info %>% select(
  ID = id, ICAO = ident, IATA = iata_code, NAME = name, LAT = latitude_deg, LON = longitude_deg
  ,COUNTRY = iso_country, REGION = iso_region ) %>%
  filter(stringr::str_detect(ICAO, "[A-Z]{4}"))

# merge with ECTRL MS
ectrl_ms <- read_csv2("./data/ECTRL_MS.csv")

apt_info <- apt_info %>% 
  left_join(
    ectrl_ms %>% select(ISO_2D, MS)
    ,by =c("COUNTRY" = "ISO_2D"))
```

add country and region to ADEP and ADES

```{r}
apt_ctry_reg <- apt_info %>% select(ICAO, COUNTRY, REGION, ECTRL = MS)

ds <- ds %>% left_join(apt_ctry_reg, by = c("ADEP" = "ICAO"))

#missing   <- ds %>% filter(is.na(COUNTRY))
#write_csv(missing, "./data/missing_apt_iso3661.csv")

ds <- ds %>% rename(ADEP_CTRY = COUNTRY, ADEP_REG = REGION, ADEP_ECTRL = ECTRL)

ds <- ds %>% left_join(apt_ctry_reg, by = c("ADES" = "ICAO")) %>%
  rename(ADES_CTRY = COUNTRY, ADES_REG = REGION, ADES_ECTRL = ECTRL)
```

Regional flight pattern

```{r}
# utility function for negated %in%
"%!in%" <- function(a,b){ ! a %in% b }

ds <- ds %>% mutate(REG_PAT = case_when(
   ADEP_ECTRL %in% c("MS")  & ADES_ECTRL %in% c("MS")  ~ "INTRA"  # TBD Israel and Morocco --> CAS
  ,ADEP_ECTRL %!in% c("MS") & ADES_ECTRL %!in% c("MS") ~ "OVR"
  ,ADEP_ECTRL %in% c("MS")  & ADES_ECTRL %!in% c("MS") ~ "OUT"
  ,ADEP_ECTRL %!in% c("MS") & ADES_ECTRL %in% c("MS")  ~ "INB"
  ,TRUE ~ "NA"
))
```



General traffic stats

```{r}
tfc <- ds %>% mutate(DOF = lubridate::as_date(LOBT)) %>% 
  group_by(DOF) %>%
  summarise(FLIGHTS = n()
            ,INTRA = sum(REG_PAT == "INTRA")
            ,OVR   = sum(REG_PAT == "OVR")
            ,OUT   = sum(REG_PAT == "OUT")
            ,INB   = sum(REG_PAT == "INB")
            ) %>%
  mutate(
    FLIGHTS_ROLL = zoo::rollmean(FLIGHTS, k = 7, fill=NA)
    ,INTRA_ROLL  = zoo::rollmean(INTRA,   k = 7, fill=NA)
    ,OVR_ROLL    = zoo::rollmean(OVR,     k = 7, fill=NA)
    ,OUT_ROLL    = zoo::rollmean(OUT,     k = 7, fill=NA)
    ,INB_ROLL    = zoo::rollmean(INB,     k = 7, fill=NA)
         )
```


```{r}
tfc %>% 
  plot_ly(x = ~DOF, y = ~FLIGHTS, name = "total flights", type = "scatter") %>%
  add_lines(y = ~ FLIGHTS_ROLL, name = "total flights (rolling average)") %>%
  add_trace(y = ~INTRA, name = "intra EUROCONTROL region") %>%
  add_lines(y = ~INTRA_ROLL, name = "intra ECTRL (rolling average)") %>%
  add_trace(y = ~OVR, name = "overflights") %>%
  add_lines(y = ~OVR_ROLL, name = "overflights (rolling average)") %>%
  add_trace(y = ~OUT, name = "out of EUROCONTROL region") %>%
  add_lines(y = ~OUT_ROLL, name = "out of ECTRL (rolling average)") %>%
  add_trace(y = ~INB, name = "inbound EUROCONTROL region") %>%
  add_lines(y = ~INB_ROLL, name = "inbound ECTRL (rolling average)") %>%
  layout(xaxis = list(title=""), yaxis = list(title = "number of flights"#, type = "log"
                                              ))
```

write out day-by-day data - initial batch

```{r}
read_NM_flight_table <- function(.fn){
  ds <- readxl::read_excel(.fn, sheet = "Sheet 1", col_names = TRUE)
}

trim_NM_flight_table <- function(.ds){
  ds <- .ds %>%  
    select(
       FLTID = AIRCRAFT_ID, REG = REGISTRATION, ICAO24 = "AIRCRAFT_ADDRESS"
      , TYPE = AIRCRAFT_TYPE_ICAO_ID, OPR = AIRCRAFT_OPERATOR
      ,"ADEP", "ADES", "LOBT", "AOBT_3", "ARVT_3"
      )
}

write_day_by_day <- function(.ds){
  ds <- .ds %>% 
    dplyr::mutate(DOF = lubridate::as_date(LOBT)) %>% 
    dplyr::group_split(DOF)
  
  fns <- ds %>% map(.f = ~paste0(unique(.$DOF),"_NM_FLT.csv"))
  
  map2( .x = ds, .y = fns
       ,.f = ~write_csv( x    = .x %>% select(-DOF)
                        ,path = paste0("./data/", .y))
       )
  return(ds %>% select(-DOF))
}
```

```{r}
fn <- "./data-src/T_FLIGHT 01012020-15012020.xlsx"

rq <- fn %>% read_NM_flight_table() %>% 
  trim_NM_flight_table() %>%
  write_day_by_day()
```

