---
title: "COVID19"
author: "RQ"
date: "22/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)
library(plotly)
library(crosstalk)
```

# Impact of COVID19 on European Air Transport

The corona virus pandemic (COVID19) is affecting air transportation around the world.
This page focusses on the European context. Refernces to global impacts/scenarios will be provided with a view to compare the response to the air traffic reduction measures been put in place by different countries.

Please consult the data page to learn more about the data behind this page.
The data is openly available via: <<link>> or is extracted from other open data projects/servers.

```{r load_data, message=FALSE}
# read in NM flight table stored in sub-folder ./data/
daily_tfcs <- list.files(path = "./data/", pattern = "*_NM_FLT.csv")
ds <- daily_tfcs %>% 
  purrr::map_dfr(.f = ~ readr::read_csv(paste0("./data/", .x)))

# read in airport refernce data downloaded from ourairports.com
apt_meta <- readr::read_csv("./data/apts_ourairports.csv") %>%
  select( ICAO = ident, IATA = iata_code
         ,COUNTRY = iso_country
         ) %>%
  filter(stringr::str_detect(ICAO, "[A-Z]{4}"))

# merge with ECTRL MS
ectrl_ms <- read_csv2("./data/ECTRL_MS.csv")

apt_meta <- apt_meta %>% 
  left_join(ectrl_ms %>% select(ISO_2D, MS), by =c("COUNTRY" = "ISO_2D"))

apt_ctry <- apt_meta %>% select(ICAO, COUNTRY, ECTRL = MS)
```

The data presented covers the 

* air traffic in Europe for the period `r daily_tfcs[1] %>% stringr::str_sub(1,10) %>% lubridate::ymd() %>% format("%d. %B %Y")` to `r daily_tfcs[length(daily_tfcs)] %>% stringr::str_sub(1,10) %>% lubridate::ymd() %>% format("%d. %B %Y")`.


```{r}
ds <- ds %>% left_join(apt_ctry, by = c("ADEP" = "ICAO")) %>%
  rename(ADEP_CTRY = COUNTRY, ADEP_ECTRL = ECTRL)

ds <- ds %>% left_join(apt_ctry, by = c("ADES" = "ICAO")) %>%
  rename(ADES_CTRY = COUNTRY, ADES_ECTRL = ECTRL)
```

```{r}
# utility function for negated %in%
"%!in%" <- function(a,b){ ! a %in% b }

ds <- ds %>% mutate(REG_PAT = case_when(
   ADEP_ECTRL %in% c("MS")  & ADES_ECTRL %in% c("MS")  ~ "INTRA"  # TBD Israel and Morocco --> CAS
  ,ADEP_ECTRL %!in% c("MS") & ADES_ECTRL %!in% c("MS") ~ "OVR"
  ,ADEP_ECTRL %in% c("MS")  & ADES_ECTRL %!in% c("MS") ~ "OUT"
  ,ADEP_ECTRL %!in% c("MS") & ADES_ECTRL %in% c("MS")  ~ "INB"
  ,TRUE ~ "NA"
))
```

```{r}
tfc <- ds %>% mutate(DOF = lubridate::as_date(LOBT)) %>% 
  group_by(DOF) %>%
  summarise(FLIGHTS = n()
            ,INTRA = sum(REG_PAT == "INTRA")
            ,OVR   = sum(REG_PAT == "OVR")
            ,OUT   = sum(REG_PAT == "OUT")
            ,INB   = sum(REG_PAT == "INB")
            ) %>%
  mutate(
    FLIGHTS_ROLL = zoo::rollmean(FLIGHTS, k = 7, fill=NA)
    ,INTRA_ROLL  = zoo::rollmean(INTRA,   k = 7, fill=NA)
    ,OVR_ROLL    = zoo::rollmean(OVR,     k = 7, fill=NA)
    ,OUT_ROLL    = zoo::rollmean(OUT,     k = 7, fill=NA)
    ,INB_ROLL    = zoo::rollmean(INB,     k = 7, fill=NA)
         )

tfc %>% 
  plot_ly(x = ~DOF, y = ~FLIGHTS, name = "total flights", type = "scatter") %>%
  add_lines(y = ~ FLIGHTS_ROLL, name = "total flights (rolling average)") %>%
  add_trace(y = ~INTRA, name = "intra EUROCONTROL region") %>%
  add_lines(y = ~INTRA_ROLL, name = "intra ECTRL (rolling average)") %>%
  add_trace(y = ~OVR, name = "overflights") %>%
  add_lines(y = ~OVR_ROLL, name = "overflights (rolling average)") %>%
  add_trace(y = ~OUT, name = "out of EUROCONTROL region") %>%
  add_lines(y = ~OUT_ROLL, name = "out of ECTRL (rolling average)") %>%
  add_trace(y = ~INB, name = "inbound EUROCONTROL region") %>%
  add_lines(y = ~INB_ROLL, name = "inbound ECTRL (rolling average)") %>%
  layout( xaxis = list(title="")
         ,yaxis = list(title = "number of flights")  #, type = "log"  
         )
```

Traffic Development at European Airports

```{r}
apt_tfc <- ds %>% 
  filter(REG_PAT %in% c("INTRA", "INB", "OUT")) %>%
  mutate(DOF = lubridate::as_date(LOBT)) 

arrs <- apt_tfc %>% filter(!is.na(ADES_ECTRL)) %>% 
  group_by(ADES, DOF) %>% 
  summarise( ARR_N = n()
            ,INTRA = sum(REG_PAT == "INTRA")
            ,OUT   = sum(REG_PAT == "OUT")
            ,INB   = sum(REG_PAT == "INB")
            ) %>%
  ungroup()
```


```{r}
top_x   <- arrs %>% group_by(ADES) %>% 
  summarise(N = n(), ARR_N = sum(ARR_N, na.rm = TRUE)) %>% 
  filter(ARR_N >= 20000)

arrs_sd <- arrs %>% filter(ADES %in% top_x$ADES) %>%
  SharedData$new(., key = ~ADES)

parrs <- arrs_sd %>% 
  plot_ly() %>% group_by(ADES) %>% 
  add_lines(x = ~DOF, y = ~ARR_N, name = ~ADES, showlegend = FALSE) %>%
  layout( xaxis = list(title = "")
         ,yaxis = list(title = "arrivals")) %>%
  highlight("plotly_selected")

bscols(widths = c(12, 12)
  ,filter_select("icao", "filter airports", arrs_sd, ~ADES)
  ,parrs
  )
```

